\LARGE \bfseries MG7: Configurable and scalable 16S metagenomics data analysis

### 2015-09-27 docs/mg7-preprint/out/paper.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.                                                 %
%                                                                                                                                                         %
% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Version 3.1 Generated 2015/22/05 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%

\documentclass{frontiersSCNS} % for Science, Engineering and Humanities and Social Sciences articles
%\documentclass{frontiersHLTH} % for Health articles
%\documentclass{frontiersFPHY} % for Physics and Applied Mathematics and Statistics articles

%\setcitestyle{square}
\usepackage{url,hyperref,lineno,microtype}
\usepackage[onehalfspacing]{setspace}
\linenumbers

% fix for pandoc 1.14
\providecommand{\tightlist}{%
\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Leave a blank line between paragraphs instead of using \\

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Alexey Alekhin {et~al.}} %use et al only if is more than 1 author
\def\Authors{
 Alexey Alekhin\(^\dagger\)\,$^{1}$  Evdokim Kovach\(^\dagger\)\,$^{1}$  Marina Manrique\,$^{1}$  Pablo Pareja-Tobes\,$^{1}$  Eduardo Pareja\,$^{1}$  Raquel Tobes\,$^{1}$  and Eduardo Pareja-Tobes $^{1,*}$
}% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$Oh no sequences! Research Group, Era7 Bioinformatics, Granada, Spain}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Eduardo Pareja-Tobes}
\def\corrAddress{Oh no sequences! Research Group, Era7 Bioinformatics, Plaza Campo Verde 3, Granada, 18001, Spain}
\def\corrEmail{eparejatobes@ohnosequences.com}

\begin{document}
\onecolumn
\firstpage{1}

\title[MG7: Configurable and scalable 16S metagenomics data analysis]{MG7: Configurable and scalable 16S metagenomics data analysis -- new
methods optimized for massive cloud computing}

\author[\firstAuthorLast ]{\Authors} %This field will be automatically populated
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}

@@

\begin{abstract}

As part of the Cambrian explosion of omics data, metagenomics brings to
the table a specific, defining trait: its social essence. The
\emph{meta} prefix exerts its influence, with multitudes manifesting
themselves everywhere; from samples to data analysis, from actors
involved to (present and future) applications. Of these dimensions, data
analysis is where needs lay further from what current tools provide. Key
features are, among others, scalability, reproducibility, data
provenance and distribution, process identity and versioning. These are
the goals guiding our work in MG7, a 16S metagenomics data analysis
system. The basic principle is a new approach to data analysis, where
configuration, processes, or data locations are static, type-checked and
subject to the standard evolution of a well-maintained software project.
Cloud computing, in its Amazon Web Services incarnation, when coupled
with these ideas, produces a robust, safely configurable, scalable tool.
Processes, data, machine behaviors and their dependencies are expressed
using a set of libraries which bring as much as possible checking and
validation to the type level, without sacrificing expressiveness.
Together they form a toolkit for defining scalable cloud-based workflows
composed of stateless computations, with a static reproducible
specification of dependencies, behavior and wiring of all steps. The
modeling of taxonomy data is done using Bio4j, where the new paradigm of
graph databases allows for both a simple expression of taxonomic
assignment tasks and the calculation of taxa abundance values
considering the hierarchic structure of the taxonomy tree. MG7 includes
a new 16S reference database, \emph{16S-DB7}, built with a flexible and
sustainable update system, and the possibility of project-driven
personalization.

\(^\dagger\) The first and second authors contributed equally to this
work.

\tiny
 \keyFont{ \section{Keywords:} Metagenomics, 16S, Bacterial diversity profile, Bio4j, Graph databases,
Cloud computing, NGS, Genomic big data, Microbiome, Environmental, 16S
Database } %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
\end{abstract}

@@

During the past decade, metagenomics data analysis is growing
exponentially. Some of the reasons behind this are the increasing
throughput of massively parallel sequencing technologies (with the
derived decrease in sequencing costs), and the wide impact of
metagenomics studies \citep{oulas2015metagenomics}, especially in human
health (diagnostics, treatments, drug response or prevention)
\citep{bikel2015combining}. We should also mention what could be called
the microbiome explosion: all kind of microbiomes (gut, mouth, skin,
urinary tract, airway, milk, bladder) are now routinely sequenced in
different conditions of health and disease, or after different
treatments. The impact of Metagenomics is also being felt in
environmental sciences \citep{ufarte2015discovery}, crop sciences, the
agrifood sector \citep{coughlan2015biotechnological} and biotechnology
in general \citep[\citet{kodzius2015marine}]{cowan2015metagenomics}.
These new possibilities for exploring the diversity of micro-organisms
in the most varied environments are opening new research areas, and
drastically changing the existing ones.

As a consequence, the challenge is thus moving (as in other fields) from
data acquisition to data analysis: the amount of data is expected to be
overwhelming in a very short time \citep{stephens2015big}.

Genome researchers have raised the alarm over big data in the past
\citep{hayden2015genome}, but even a more serious challenge might be
faced with the metagenomics boom. If we compare metagenomics data with
other genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling from the same patient
\citep{faust2015metagenomics} along several weeks (or years) is being
used for the follow up of some intestinal pathologies, for studying the
evolution of the gut microbiome after antibiotic treatment, or for colon
cancer early detection
\citep[\citet{garrett2015cancer}]{zeller2014potential}. This need of
sampling across time adds more complexity to metagenomics data storage
and demands adapted algorithms to detect state variations across time as
well as idiosyncratic commonalities of the microbiome of each individual
\citep{franzosa2015identifying}. In addition to the intra-individual
sampling-time dependence, metagenomic clinical test results vary
depending on the specific region of extraction of the clinical specimen.
This local variability adds complexity to the analysis since different
localizations (different tissues, different anatomical regions, healthy
or tumor tissues) are required to have a sufficiently complete landscape
of the human microbiome. Moreover, re-analysis of old samples using new
tools and better reference databases might be also demanded from time to
time.

Other disciplines such as astronomy or particle physics have faced the
big data challenge before. A key difference is the existence of
standards for data processing \citep{stephens2015big}; in metagenomics
global standards for converting raw sequence data into processed data
are not yet well defined, and there are shortcomings derived from the
fact that most bioinformatics methodologies used for metagenomics data
analysis were designed for scenarios very different from the current
one. These are some of the aspects that have suffered crucial changes
and advances with a direct impact in metagenomics data analysis:

@@

\href{https://github.com/ohnosequences/datasets}{Datasets} is a Scala
library developed by the first and last authors with the goal of being a
Scala-embedded mini-language for datasets and their locations.
\textbf{Data} is represented as type-indexed fields: keys are modeled as
singleton types, and values correspond to what could be called a
denotation of the key: a value of type \texttt{Location} tagged with the
key type. Then a \textbf{Dataset} is essentially a collection of data,
which are guaranteed statically to be different through type-level
predicates, making use of the value--type correspondence which can be
established through singleton types and implicits. A dataset location is
then just a list of locations formed by locations of each dataset key.
All this is based on what could be described as an embedding in Scala of
an extensible record system with concatenation on disjoint labels, in
the spirit of \citep[\citet{harper1991record}]{harper1990extensible}.
For that \emph{Datasets} uses
\href{https://github.com/ohnosequences/cosas/}{ohnosequences/cosas}
library.

@@

We developed our own Scala-based type-safe eDSLs (embedded Domain
Specific Languages) for
\href{https://github.com/ohnosequences/flash}{FLASh}
\citep{magovc2011flash} and
\href{https://github.com/ohnosequences/blast}{BLAST}
\citep{camacho2009blast} expressions and their execution.

@@

\citep[Bio4j][]{pareja2015bio4j} is a data platform integrating data
from different resources such as UniProt or GO in a graph data paradigm.
In the assignment phase we use a subgraph containing the NCBI Taxonomy,
wrapping in Scala its Java API in a tree algebraic data type.

@@

Our 16S-DB7 Reference Database is a curated subset of sequences from the
NCBI nucleotide database \textbf{nt}. The sequences included were
selected by similarity with the bacterial and archaeal reference
sequences downloaded from the \textbf{RDP database}
\citep{cole2013ribosomal}. RDP unaligned sequences were used to capture
new 16S RNA sequences from \textbf{nt} using BLAST similarity search
strategies and then, performing additional curation steps to remove
sequences with poor taxonomic assignments to taxonomic nodes close to
the root of the taxonomy tree. All the nucleotide sequences included in
\textbf{nt} database has a taxonomic assignment provided by the
\textbf{Genbank} sequence submitter. NCBI provides a table (available at
ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/) to do the mapping of any
Genbank Identifier (GI) to its Taxonomy Identifier (TaxID). Thus, we are
based on a crowdsourced submitter-maintained taxonomic annotation system
for reference sequences. It supposes a sustainable system able to face
the expected number of reference sequences that will populate the public
global nucleotide databases in the near future. Another advantageous
point is that we are based on NCBI taxonomy, the \emph{de-facto}
standard taxonomic classification for biomolecular data
\citep{cochrane20102010}. NCBI taxonomy is, undoubtedly, the most used
taxonomy all over the world and the most similar to the official
taxonomies of each specific field. This is a crucial point because all
the type-culture and tissue databanks follow this official taxonomical
classification and, in addition, all the knowledge accumulated during
last decades is referred to this taxonomy. In addition NCBI provides a
direct connection between taxonomical formal names and the physical
specimens that serve as exemplars for the species
\citep{federhen2014type}.

@@

@@

In the first step the paired-end reads, designed with an insert size
that yields pairs of reads with an overlapping region between them, are
assembled using FLASh \citep{magovc2011flash}. FLASh is designed to
merge pairs of reads when the original DNA fragments are shorter than
twice the length of reads. Thus, the sequence obtained after joining the
2 reads of each pair is larger and has better quality since the sequence
at the ends of the reads is refined merging both ends in the assembly.
To have a larger and improved sequence is crucial to do more precise the
inference of the bacterial origin based on similarity with reference
sequences.

@@

The second step is to search for similar 16S sequences in our 16S-DB7
database. The taxonomic assignment for each read is based on BLASTN of
each read against the 16S database.Assignment based on direct similarity
of each read one by one compared against a sufficiently wide database is
considered in different reviews of metagenomics analysis methodologies
\citep[\citet{morgan2012chapter}]{segata2013computational} as a very
exhaustive method for assignment. Some methods of assignment compare the
sequences only against the 16S genes from available complete bacterial
genomes or avoid computational cost clustering or binning the sequences
first, and then doing the assignments only for the representative
sequence of each cluster. MG7 carries out an exhaustive comparison of
all the reads under analysis and it does not applies any binning
strategy. Every read is specifically compared with all the sequences of
the 16S database. We select the best BLAST hits (10 hits by default)
obtained for each read to do the taxonomic assignment.

>>>>

% \documentclass[a4paper]{article}
\documentclass[fontsize=8pt,paper=a4,pagesize,abstracton]{scrartcl}

% to prevent funny stuff happening with old packages (I think)
\usepackage{ifxetex}
% xetex font-related packages
\usepackage{fontspec,xltxtra,xunicode}
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% I don't know why this is here
\usepackage{fixltx2e}

%%% color definitions %%%
\usepackage[usenames,dvipsnames,xetex]{xcolor}
% old cyan variants %
\definecolor{OldCyan}{RGB}{62,127,137}
\definecolor{OldCyan-Dark}{RGB}{38,104,113}
\definecolor{OldCyan-Light}{RGB}{86,156,166}
\definecolor{OldCyan-Matte}{RGB}{62,113,120}
\definecolor{OldCyan-MatteLight}{RGB}{95,158,166}
% light amber variants %
\definecolor{LightAmber}{RGB}{223,184,98}
\definecolor{LightAmber-Dark}{RGB}{184,145,61}
\definecolor{LightAmber-Light}{RGB}{231,195,118}
\definecolor{LightAmber-Matte}{RGB}{195,165,100}
\definecolor{LightAmber-MatteLight}{RGB}{231,199,131}
% salmon variants %
\definecolor{Salmon}{RGB}{223,107,98}
\definecolor{Salmon-Dark}{RGB}{184,70,61}
\definecolor{Salmon-Light}{RGB}{231,126,118}
\definecolor{Salmon-Matte}{RGB}{195,107,100}
\definecolor{Salmon-MatteLight}{RGB}{231,138,131}
\definecolor{Grey}{RGB}{88,88,88}
\definecolor{Grey-Light}{RGB}{129,129,129}
\definecolor{Grey-Dark}{RGB}{65,65,65}
% SOLARIZED
\definecolor{solarized@base03}{HTML}{002B36}
\definecolor{solarized@base02}{HTML}{073642}
\definecolor{solarized@base01}{HTML}{586e75}
\definecolor{solarized@base00}{HTML}{657b83}
\definecolor{solarized@base0}{HTML}{839496}
\definecolor{solarized@base1}{HTML}{93a1a1}
\definecolor{solarized@base2}{HTML}{EEE8D5}
\definecolor{solarized@base3}{HTML}{FDF6E3}
\definecolor{solarized@yellow}{HTML}{B58900}
\definecolor{solarized@orange}{HTML}{CB4B16}
\definecolor{solarized@red}{HTML}{DC322F}
\definecolor{solarized@magenta}{HTML}{D33682}
\definecolor{solarized@violet}{HTML}{6C71C4}
\definecolor{solarized@blue}{HTML}{268BD2}
\definecolor{solarized@cyan}{HTML}{2AA198}
\definecolor{solarized@green}{HTML}{859900}

% math symbols and such
\usepackage{amsmath}
\usepackage{unicode-math}

% english support
\usepackage{polyglossia}
\setmainlanguage{english}

% paper size and lead
\usepackage[]{geometry}
\geometry{a4paper}
\usepackage{setspace}
% default lead is 1.216666
\setstretch{1.15}

% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{parskip}

% references

% for natbib

% for biblatex
\usepackage[
  backend=biber,
  texencoding=utf8,
  bibencoding=utf8,
  hyperref=true,
  maxbibnames=15,
  isbn=false,
  dateabbrev=false,
  style=ieee
]{biblatex}

\renewcommand*{\newunitpunct}{\space}
\renewcommand{\bibfont}{\small}
\setlength\bibitemsep{5pt}

\renewbibmacro*{volume+number+eid}{%
  \printfield{volume}%
  \setunit{\,}% DELETED
  \printfield{number}%
  \setunit{.\space}%
  \printfield{eid}}

\DeclareFieldFormat[article,inbook,incollection,inproceedings,patent,thesis,unpublished]{volume}{\mkbibbold{#1}}
\DeclareFieldFormat{date}{#1}
\DeclareFieldFormat{pages}{#1\adddot}
\DeclareFieldFormat[article,inbook,incollection,inproceedings,patent,thesis,unpublished]{number}{\mkbibparens{#1}\addcolon}

\DeclareFieldFormat{doi}{\newline DOI: \href{https://dx.doi.org/#1}{#1}}
\DeclareFieldFormat{url}{}
\DeclareFieldFormat[unpublished]{url}{\newline URL: \href{#1}{#1}}
\DeclareFieldFormat
  [article,inbook,incollection,inproceedings,patent,thesis,unpublished]
  {journaltitle}{\newline\emph{#1}}

\DeclareFieldFormat
  [article,inbook,incollection,inproceedings,patent,thesis,unpublished]
  {title}{\textbf{#1}}

\bibliography{./out/../refs.bib}

\usepackage{longtable}
\usepackage{ctable}
\usepackage{float}
\usepackage[normalem]{ulem}
% typography settings
% default conf applied

% Open Sans, a beautiful fontface
\setmainfont[Mapping = tex-text, Color = Grey, BoldFont = {* Semibold}, BoldItalicFont = {* Semibold Italic}, ItalicFont = {* Italic}]{Open Sans}
% sort of like Cambria Math; somehow matches a sans body text
\setmathfont{Asana Math} % math font
% font for sections: Bitter
\newfontfamily\sectfont
  [BoldFont={* Bold},
   Numbers={OldStyle,Proportional},
   Color = Grey,
   Mapping = tex-text
  ]
  {Bitter}
% the title font
\newfontfamily\titlefont
  [BoldFont={Open Sans},
   Numbers={OldStyle,Proportional},
   Mapping = tex-text
  ]
  {Open Sans Light}

% for changing the sections font
\usepackage{sectsty}
\allsectionsfont{\sectfont\textmd}

% verbatim and code highlighting: Solarized Light
\usepackage{fancyvrb}
\DefineShortVerb[commandchars=\\\{\}]{\|}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{fontsize=\footnotesize,%
commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
% this is solarized light
\definecolor{shadecolor}{RGB}{253,246,227} % solarized@base3
\newenvironment{Shaded}{\vspace{\baselineskip}\begin{shaded}}{\end{shaded}\vspace{\baselineskip}}
% colored backgrd for verb
\let\oldverbatim=\verbatim
\let\endoldverbatim=\endverbatim
\renewenvironment{verbatim}[1]{%
  \vspace{\baselineskip}
  \scriptsize
  \par\setstretch{1}
  \begin{shaded}
  \begin{oldverbatim}{#1}%
}%
{%
  \end{oldverbatim}%
  \end{shaded}
  \vspace{\baselineskip}
}

\newcommand{\KeywordTok}[1]{\textbf{\textcolor{solarized@green}{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor{solarized@blue}{#1}}
\newcommand{\DecValTok}[1]{\textcolor{solarized@violet}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor{solarized@violet}{#1}}
\newcommand{\FloatTok}[1]{\textcolor{solarized@violet}{#1}}
\newcommand{\CharTok}[1]{\textcolor{solarized@cyan}{#1}}
\newcommand{\StringTok}[1]{\textcolor{solarized@cyan}{#1}}
\newcommand{\CommentTok}[1]{\textcolor{solarized@base1}{\textit{#1}}}
\newcommand{\OtherTok}[1]{\textcolor{solarized@base00}{#1}}
\newcommand{\AlertTok}[1]{\textcolor{solarized@yellow}{\textbf{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor{solarized@blue}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor{solarized@base1}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor{solarized@red}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{\textcolor{solarized@base00}{#1}}

% monospace also grey; Source Code Pro
\setmonofont[Mapping=tex-ansi,Color=solarized@base00,Numbers=Lining,Ligatures=NoRequired]{Source Code Pro}

% allow to break lines more easily on tt text
% http://tex.stackexchange.com/questions/52850/temporarily-increase-the-limit-on-space-size
\let\OldTexttt\texttt
\renewcommand{\texttt}[1]{ \emergencystretch=2em \OldTexttt{#1} }
% maybe?
\setlength{\emergencystretch}{3em}  % prevent overfull lines

% color fixes
\makeatletter
\newcommand{\globalcolor}[1]{%
  \color{#1}\global\let\default@color\current@color
}
\makeatother

\AtBeginDocument{\globalcolor{Grey}}
\renewcommand{\textcolor}[2]{{\addfontfeature{Color=#1}#2}}

% Lists formatting:
% note sure what 'fancy enums' are; something to do with lists,
% as the further comment suggests:

% tables
\usepackage{tabu}
\taburulecolor{Salmon-Dark}
\let\quoteOld\quote
\let\endquoteOld\endquote
\renewenvironment{quote}{\vspace{\baselineskip}\itshape\quoteOld\begin{tabu}{|[2pt]p{.9\textwidth}}}{\end{tabu}\endquoteOld\vspace{\baselineskip}}

% longtable needed by latest pandoc version
\usepackage{longtable}

\setupctable{
captionskip=0pt, framerule=0pt, nostar,
center, framesep=0pt, pos=tbp,
continued=(continued), maxwidth=0pt, super,
doinside={}, mincapwidth=0pt, table,
framebg=1 1 1, nonotespar, topcap,
framefg=.3 .1 .68, nosideways, width=0pt
}

% math diagrams
\usepackage[all]{xy}
\xyoption{pdf}  % improves rendering

\usepackage[breaklinks=true]{hyperref}
\hypersetup{
  setpagesize=false, % page size defined by xetex
  unicode=false, % unicode breaks when used with xetex
  xetex,
  pdfnewwindow,
  colorlinks,%
  citecolor=OldCyan,%
  filecolor=OldCyan,%
  linkcolor=OldCyan,%
  urlcolor=OldCyan,
  urlbordercolor=OldCyan
}

\let\OldHref\href
\renewcommand{\href}[2]{\OldHref[pdfnewwindow]{#1}{{#2}}}

% ugly fix for latest pandoc
\providecommand{\tightlist}{%
\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% ugly fix for link colors in xetex
\makeatletter
\def\HyColor@@@@UseColor#1\@nil{\addfontfeatures{Color=#1}}
\makeatother

% images
\usepackage{graphicx}
% -- We will generate all images so they have a width \maxwidth. This means
% -- that they will get their normal width if they fit onto the page, but
% -- are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}

% this is wrong
\setcounter{secnumdepth}{0}

  \title{\LARGE \bfseries MG7: Configurable and scalable 16S metagenomics data analysis}

% authors
\usepackage{authblk}
  \author[1]{\normalsize Alexey Alekhin\(^\dagger\)}
  \author[1]{\normalsize Evdokim Kovach\(^\dagger\)}
  \author[1]{\normalsize Marina Manrique}
  \author[1]{\normalsize Pablo Pareja-Tobes}
  \author[1]{\normalsize Eduardo Pareja}
  \author[1]{\normalsize Raquel Tobes}
  \author[1]{\normalsize Eduardo Pareja-Tobes}

\affil[1]{\small
  {\itshape\href{http://ohnosequences.com}{oh no sequences!}} research group, \href{http://www.era7bioinformatics.com}{Era7 bioinformatics}
}

\date{\normalsize \today}

% page style!
% spaces
\setlength{\footskip}{3\baselineskip}
\setlength{\headsep}{3\baselineskip}
\setlength{\skip\footins}{0cm}
% \setlength{\voffset}{-10pt}
\setlength{\topmargin}{0pt}
% \setlength{\textheight}{612pt}

\makeatletter
\let\Title\@title
\let\Author\@author
\makeatother

\usepackage{fancyhdr}
\pagestyle{fancy}

\renewcommand{\sectionmark}[1]{%
\markboth{#1}{}}
% \fancyheadoffset{0cm}
\renewcommand{\headrulewidth}{.5pt}
\renewcommand{\headrule}{\vbox to 0pt{\hbox
to\headwidth{\dotfill}\vss}}

\lhead{\titlefont MG7: Configurable and scalable 16S metagenomics data analysis}
% \chead{\titlefont {\color{Grey-Light}\leftmark}}
\chead{}
\rhead{\titlefont \thepage}

\lfoot{\titlefont {{\color{Salmon-Dark}oh}{\color{LightAmber-Dark}no}{\color{Grey}sequences}{\color{Salmon-Dark}!} }}
\cfoot{}
\rfoot{\titlefont {\color{Grey-Light} era7 bioinformatics R\&D group}}
% footer rule
\renewcommand{\footruleskip}{0.66\baselineskip}
\renewcommand{\footrulewidth}{.5pt}
\renewcommand{\footrule}{{\color{LightAmber-MatteLight}%
\vskip-\footruleskip\vskip-\footrulewidth
\hrule width\headwidth height\footrulewidth\vskip\footruleskip}}

\renewcommand{\footrule}{\vbox to 0pt{\hbox
to\headwidth{\color{Salmon}{\dotfill}}\vss}}

\begin{document}

@@

\begin{abstract} % \footnotesize
As part of the Cambrian explosion of omics data, metagenomics brings to
the table a specific, defining trait: its social essence. The
\emph{meta} prefix exerts its influence, with multitudes manifesting
themselves everywhere; from samples to data analysis, from actors
involved to (present and future) applications. Of these dimensions, data
analysis is where needs lay further from what current tools provide. Key
features are, among others, scalability, reproducibility, data
provenance and distribution, process identity and versioning. These are
the goals guiding our work in MG7, a 16S metagenomics data analysis
system. The basic principle is a new approach to data analysis, where
configuration, processes, or data locations are static, type-checked and
subject to the standard evolution of a well-maintained software project.
Cloud computing, in its Amazon Web Services incarnation, when coupled
with these ideas, produces a robust, safely configurable, scalable tool.
Processes, data, machine behaviors and their dependencies are expressed
using a set of libraries which bring as much as possible checking and
validation to the type level, without sacrificing expressiveness.
Together they form a toolkit for defining scalable cloud-based workflows
composed of stateless computations, with a static reproducible
specification of dependencies, behavior and wiring of all steps. The
modeling of taxonomy data is done using Bio4j, where the new paradigm of
graph databases allows for both a simple expression of taxonomic
assignment tasks and the calculation of taxa abundance values
considering the hierarchic structure of the taxonomy tree. MG7 includes
a new 16S reference database, \emph{16S-DB7}, built with a flexible and
sustainable update system, and the possibility of project-driven
personalization. \newline\newline
\(^\dagger\) The first and second authors contributed equally to this
work.
\end{abstract}

\thispagestyle{empty}

% \setcounter{tocdepth}{3}
\tableofcontents

\newpage

@@

During the past decade, metagenomics data analysis is growing
exponentially. Some of the reasons behind this are the increasing
throughput of massively parallel sequencing technologies (with the
derived decrease in sequencing costs), and the wide impact of
metagenomics studies \autocite{oulas2015metagenomics}, especially in
human health (diagnostics, treatments, drug response or prevention)
\autocite{bikel2015combining}. We should also mention what could be
called the microbiome explosion: all kind of microbiomes (gut, mouth,
skin, urinary tract, airway, milk, bladder) are now routinely sequenced
in different conditions of health and disease, or after different
treatments. The impact of Metagenomics is also being felt in
environmental sciences \autocite{ufarte2015discovery}, crop sciences,
the agrifood sector \autocite{coughlan2015biotechnological} and
biotechnology in general \autocite{cowan2015metagenomics}
\autocite{kodzius2015marine}. These new possibilities for exploring the
diversity of micro-organisms in the most varied environments are opening
new research areas, and drastically changing the existing ones.

As a consequence, the challenge is thus moving (as in other fields) from
data acquisition to data analysis: the amount of data is expected to be
overwhelming in a very short time \autocite{stephens2015big}.

Genome researchers have raised the alarm over big data in the past
\autocite{hayden2015genome}, but even a more serious challenge might be
faced with the metagenomics boom. If we compare metagenomics data with
other genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling from the same patient
\autocite{faust2015metagenomics} along several weeks (or years) is being
used for the follow up of some intestinal pathologies, for studying the
evolution of the gut microbiome after antibiotic treatment, or for colon
cancer early detection \autocite{zeller2014potential}
\autocite{garrett2015cancer}. This need of sampling across time adds
more complexity to metagenomics data storage and demands adapted
algorithms to detect state variations across time as well as
idiosyncratic commonalities of the microbiome of each individual
\autocite{franzosa2015identifying}. In addition to the intra-individual
sampling-time dependence, metagenomic clinical test results vary
depending on the specific region of extraction of the clinical specimen.
This local variability adds complexity to the analysis since different
localizations (different tissues, different anatomical regions, healthy
or tumor tissues) are required to have a sufficiently complete landscape
of the human microbiome. Moreover, re-analysis of old samples using new
tools and better reference databases might be also demanded from time to
time.

Other disciplines such as astronomy or particle physics have faced the
big data challenge before. A key difference is the existence of
standards for data processing \autocite{stephens2015big}; in
metagenomics global standards for converting raw sequence data into
processed data are not yet well defined, and there are shortcomings
derived from the fact that most bioinformatics methodologies used for
metagenomics data analysis were designed for scenarios very different
from the current one. These are some of the aspects that have suffered
crucial changes and advances with a direct impact in metagenomics data
analysis:

@@

\href{https://github.com/ohnosequences/datasets}{Datasets} is a Scala
library developed by the first and last authors with the goal of being a
Scala-embedded mini-language for datasets and their locations.
\textbf{Data} is represented as type-indexed fields: keys are modeled as
singleton types, and values correspond to what could be called a
denotation of the key: a value of type \texttt{Location} tagged with the
key type. Then a \textbf{Dataset} is essentially a collection of data,
which are guaranteed statically to be different through type-level
predicates, making use of the value--type correspondence which can be
established through singleton types and implicits. A dataset location is
then just a list of locations formed by locations of each dataset key.
All this is based on what could be described as an embedding in Scala of
an extensible record system with concatenation on disjoint labels, in
the spirit of \autocite{harper1990extensible}
\autocite{harper1991record}. For that \emph{Datasets} uses
\href{https://github.com/ohnosequences/cosas/}{ohnosequences/cosas}
library.

@@

We developed our own Scala-based type-safe eDSLs (embedded Domain
Specific Languages) for
\href{https://github.com/ohnosequences/flash}{FLASh}
\autocite{magovc2011flash} and
\href{https://github.com/ohnosequences/blast}{BLAST}
\autocite{camacho2009blast} expressions and their execution.

@@

Bio4j \autocite{pareja2015bio4j} is a data platform integrating data
from different resources such as UniProt or GO in a graph data paradigm.
In the assignment phase we use a subgraph containing the NCBI Taxonomy,
wrapping in Scala its Java API in a tree algebraic data type.

@@

Our 16S-DB7 Reference Database is a curated subset of sequences from the
NCBI nucleotide database \textbf{nt}. The sequences included were
selected by similarity with the bacterial and archaeal reference
sequences downloaded from the \textbf{RDP database}
\autocite{cole2013ribosomal}. RDP unaligned sequences were used to
capture new 16S RNA sequences from \textbf{nt} using BLAST similarity
search strategies and then, performing additional curation steps to
remove sequences with poor taxonomic assignments to taxonomic nodes
close to the root of the taxonomy tree. All the nucleotide sequences
included in \textbf{nt} database has a taxonomic assignment provided by
the \textbf{Genbank} sequence submitter. NCBI provides a table
(available at ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/) to do the
mapping of any Genbank Identifier (GI) to its Taxonomy Identifier
(TaxID). Thus, we are based on a crowdsourced submitter-maintained
taxonomic annotation system for reference sequences. It supposes a
sustainable system able to face the expected number of reference
sequences that will populate the public global nucleotide databases in
the near future. Another advantageous point is that we are based on NCBI
taxonomy, the \emph{de-facto} standard taxonomic classification for
biomolecular data \autocite{cochrane20102010}. NCBI taxonomy is,
undoubtedly, the most used taxonomy all over the world and the most
similar to the official taxonomies of each specific field. This is a
crucial point because all the type-culture and tissue databanks follow
this official taxonomical classification and, in addition, all the
knowledge accumulated during last decades is referred to this taxonomy.
In addition NCBI provides a direct connection between taxonomical formal
names and the physical specimens that serve as exemplars for the species
\autocite{federhen2014type}.

@@

\begin{figure}[htbp]
\centering
\includegraphics{Figure-1.jpg}
\caption{\textbf{MG7 analysis workflow.} The paired reads in fastq
format are merged resulting in only one sequence per read pair. The next
step is a parallelized BLASTN of every merged sequence against the 16S
reference database 16S-DB7. Then, the mapping of the detected similar
sequences in the database to the taxonomy node to which they belong is
carried out. This is done using Bio4j that includes a module with all
the NCBI taxonomy in a graph connected with the Gene Ontology, Uniprot,
and RefSeq graphs. Then the taxonomic assignment is done for each
sequence following two different approaches: LCA and BBH, and finally
the abundances corresponding to direct and cumulative assignments for
each node in percentage and absolute counts are provided for each
assignment mode.}
\end{figure}

@@

In the first step the paired-end reads, designed with an insert size
that yields pairs of reads with an overlapping region between them, are
assembled using FLASh \autocite{magovc2011flash}. FLASh is designed to
merge pairs of reads when the original DNA fragments are shorter than
twice the length of reads. Thus, the sequence obtained after joining the
2 reads of each pair is larger and has better quality since the sequence
at the ends of the reads is refined merging both ends in the assembly.
To have a larger and improved sequence is crucial to do more precise the
inference of the bacterial origin based on similarity with reference
sequences.

@@

The second step is to search for similar 16S sequences in our 16S-DB7
database. The taxonomic assignment for each read is based on BLASTN of
each read against the 16S database.Assignment based on direct similarity
of each read one by one compared against a sufficiently wide database is
considered in different reviews of metagenomics analysis methodologies
\autocite{segata2013computational} \autocite{morgan2012chapter} as a
very exhaustive method for assignment. Some methods of assignment
compare the sequences only against the 16S genes from available complete
bacterial genomes or avoid computational cost clustering or binning the
sequences first, and then doing the assignments only for the
representative sequence of each cluster. MG7 carries out an exhaustive
comparison of all the reads under analysis and it does not applies any
binning strategy. Every read is specifically compared with all the
sequences of the 16S database. We select the best BLAST hits (10 hits by
default) obtained for each read to do the taxonomic assignment.
